An open and continuously updated fern tree of life (FTOL)

### 2022-01-25 ms/manuscript.Rmd
Update MS

**Joel H. Nitta^1^, Eric Schuettpelz^2^, and Wataru Iwasaki^3^**

@@

^3^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan

@@

@@

Spacer regions are difficult to align across higher taxonomic levels within ferns (e.g., family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer levels (e.g., within genera) where slower-evolving genes like *rbcL* may not provide enough resolution.
Therefore, we first aligned spacer regions for each family with MAFFT, then merged the subalignments using the MAFFT "--merge" option.
This retains the indels within each subalignment while aligning across subalignments.
For some small families with fewer than three species each that could be used for subalignments, these were added as "singletons" during the MAFFT "--merge".

>>>>

**Joel H. Nitta^1^, Eric Schuettpelz^2^, Santiago Ramírez-Barahona^3^, and Wataru Iwasaki^4^**

@@

^3^ Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Mexico City, Mexico

^4^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan

@@

Ferns (ca. 12,000 spp.) are the second most diverse clade of vascular plants after seed plants (ca. 260,000 spp.), and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [e.g., @Hennequin2014; @Lehtonen2015], and speciation [e.g., @Kao2020].
As Dobzhansky famously said, "Nothing in biology makes sense except in the light of evolution"; likewise, a well-sampled molecular phylogeny is key to any investigation of fern evolutionary biology.
Fortunately, ferns have received relatively intense focus from molecular systematists, resulting in the publication of molecular trees for nearly all major clades and DNA sequences for ca. 1/3 of all fern species deposited in GenBank.
Thus, there is both sufficient sampling and a pressing need for a globally sampled fern phylogeny.

Past efforts to construct such a global phylogeny have steadily increased in number of species as sequences in GenBank have accumulated [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
Indeed, accumulation of plastid fern accessions in GenBank show no sign of slowing since the most recent global phylogeny [@Testo2016a].
We believe there is a need therefore, not only for a global fern phylogeny, but one that is self-updating to keep pace with the continued accumulation of molecular data.
This will eliminate the need for researchers to "rebuild the wheel" each time they want to use a globally sampled fern phylogeny.

We recognize that multiple efforts exist to automatically or semi-automatically generate trees for any arbitrary part of the tree of life [@Antonelli2016], a tree for all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015] which would subsume a global fern phylogeny.
While these may be appropriate in some cases, we contend that they cannot produce an optimal fern phylogeny due to the need to take "one-size-fits-all" approaches to accommodate such a wide phylogenetic breadth. 
By focusing methods and datasets specifically on ferns, it should be possible to generate a higher quality end-product (tree) that can be used "as-is" by biologists studying these organisms.

Furthermore, there is much to be gained from integrating a carefully designed global fern phylogeny with the fern systematics community that would not be possible with a "tree of all life" or "tree of all plants".
Recently, a genus-level taxonomy of ferns and lycophytes was established following a open, community-driven model [@PteridophytePhylogenyGroupI2016].
Thanks to this community-first approach, the taxonomy (PPG I) has been widely accepted and used (cited 475 times on Google Scholar as of writing).
However, there were clearly problematic (non-monophyletic) genera at the time PPG I was published, and many new taxonomic changes have been (and will continue to be) made since [e.g., @Almeida2017; @Shang2018; @Zhang2020].
It is anticipated that "PPG II" will be a online, open resource that can be updated as necessary (Schuettpelz, pers. comm.).
We envision an open, continuously updated global fern phylogeny that is directly integrated with PPG II so that taxonomic decisions reflecting phylogeny can be made based on community consensus using the most recently available data.

Here, we leverage our specialized taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL).
We plan to run the pipeline on a semi-annual basis and make the results freely available online through a web portal and R package.
This will enable anybody interested in the biology of ferns to have access to the most recent hypothesis of fern phylogeny.
We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) It will always provide the most up-to-date snapshot of our collective understanding of fern phylogeny; 2) It will allow for continuous assessment of taxonomy, and indicate what parts of the tree are in need of taxonomic revision; 3) It will be an important source of data for any phylogenetic comparative study of ferns.

@@

Spacer regions are difficult to align across higher taxonomic levels within ferns (e.g., family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer levels (e.g., within genera) where slower-evolving genes like *rbcL* may not provide enough resolution.
Therefore, we first aligned spacer regions for each family with MAFFT, then merged the subalignments using the MAFFT "--merge" option.
This retains the indels within each subalignment while aligning across subalignments.
For some small families with fewer than three species each that could be used for subalignments, these were added as "singletons" during the MAFFT "--merge".
Anemiaceae showed an extremely high number of indels compared to other families and could not be reliably aligned across families, so we excluded it from the spacer region data.
We also excluded outgroups from the spacer region data as they cannot be reliably aligned to ferns.
